VarCovar Rembrandt

### 2023-09-22 reporting/simu_rembrandt_breast.Rmd
update report with new simulation
Update report with new results

@@

## Beta

@@

## Variance

@@

## Correlation

@@

@@

## Summary Rembrandt

@@

- number of genes : $NG = \{10, 50, 200\}$
- proportion of significant genes : $p_g = \{0.2, 0.5\}$
- Variance : $C_{jj} = 0.2$
- Type (D) : $\beta \sim \mathcal{N}(0, 0.4^2)$
- Type (E) : half of genes follows $\beta \sim \mathcal{N}(-0.4, 0.2^2)$ and half follows $\beta \sim \mathcal{N}(0.4, 0.2^2)$
- Type (F) : half of genes follows $\beta \sim \mathcal{N}(-0.8, 0.4^2)$ and half follows $\beta \sim \mathcal{N}(0.8, 0.4^2)$
- Case (IV) : Overall correlation follows a non-standard beta $corr \sim NSBeta(20, 20, min=-1, max =1)$
- Case (V) : Correlation between significant genes $corr \sim NSBeta(10, 10, min=-1, max =1)$ and between other genes $corr \sim NSBeta(25, 25, min=-1, max =1)$

Nb of simulations to perform = $3*2*3*2 = 36$

@@

```{r fig.height=10, fig.cap="Statistical power depending on simulation setting. N is the number of patients and NG the number of genes in the pathway."}
source("script/script_simulation/figures_simu_rembrandt_breast_v1.R")

@@

Overall, all methods have close statistical power but sGBJ and Global boost test tends to be among the worst methods. There are some exceptions to note :

- Global Boost Test performs well in type F
- sGBJ performs well when N = 100 and NG = 50 which might be related to its asymptotic nature.
- In other cases, Wald and Global Test slightly outperformed other methods and Wald seems better than Global Test.

@@

- estimate one beta by gene using cox model
- sum the wald test statistics
- H0 : variance of beta is equal to 0.

@@

sGBJ is similar to Wald and Global test in the sense that it is based on one cox model by feature. However, it takes into account the covariance of genes in its formula. And the data are simulated with no covariance of beta => Problem ? look at GBJ paper

>>>>

We used Rembrandt and breast cancer datasets. For each datasets we explored the parameters of interest to perform the simulation study.

@@

## Beta from Cox model

@@

## Gene expression variance

@@

## Gene expression correlation

### Overall gene expression correlation

@@

### Gene expression correlation depending on gene significance

@@

## Summary Rembrandt and simulation settings

@@

- number of genes : $NG = \{10, 50, 200\}$
- number of individuals : $N = \{50, 100\}$
- proportion of significant genes : $p_g = \{0.2, 0.5\}$
- proportion of censored genes : $c = \{0, 0.3\}$
- Variance : $C_{jj} = 0.2$

<!-- **Note : scenario D, E and F do not rely on correlation matrix to simulate variance covariance matrix so the those matrix are unlikely to be positive definite.** -->

### Case

- Case 4 : Overall correlation follows a non-standard beta $corr \sim NSBeta(20, 20, min=-1, max =1)$
- Case 5 : Correlation between significant genes $corr \sim NSBeta(10, 10, min=-1, max =1)$ and between other genes $corr \sim NSBeta(25, 25, min=-1, max =1)$

**Note : Case IV and V are not guaranteed to generate positive definite matrix, after the correlation values are sampled, if the matrix is not positive definite, the nearest positive definite matrix is computed using the algorithm defined by Higham 2002 (see Matrix::nearPD). This correction was not implemented for scenario D, E and F.**

- Case 6 : Correlation between significant genes is $0.2$, correlation with non significant genes is $0$.

### Type

- Type (D) : $\beta \sim \mathcal{N}(0, 0.4^2)$
- Type (E) : half of genes follows $\beta \sim \mathcal{N}(-0.4, 0.2^2)$ and half follows $\beta \sim \mathcal{N}(0.4, 0.2^2)$
- Type (F) : half of genes follows $\beta \sim \mathcal{N}(-0.8, 0.4^2)$ and half follows $\beta \sim \mathcal{N}(0.8, 0.4^2)$
- Type (G) : $\beta \sim \mathcal{N}(0, 0.4^2)$. Correlation matrix of $\beta$ is the same as correlation matrix of genes.
- Type (H) : half of genes follows $\beta \sim \mathcal{N}(-0.4, 0.2^2)$ and half follows $\beta \sim \mathcal{N}(0.4, 0.2^2)$. Correlation matrix of $\beta$ is the same as correlation matrix of genes.
- Type (I) : half of genes follows $\beta \sim \mathcal{N}(-0.8, 0.4^2)$ and half follows $\beta \sim \mathcal{N}(0.8, 0.4^2)$. Correlation matrix of $\beta$ is the same as correlation matrix of genes.

@@

```{r fig.height=10, fig.cap="Statistical power depending on simulation setting. N is the number of patients and NG the number of genes in the pathway. Note : scenario D, E and F do not rely on correlation matrix to simulate variance covariance matrix so the those matrix are unlikely to be positive definite."}
source("script/script_simulation/figures_simu_rembrandt_breast_v1.R")

@@

```{r fig.height=14, fig.cap="Statistical power depending on simulation setting (second batch). N is the number of patients and NG the number of genes in the pathway."}
source("script/script_simulation/figures_simu_rembrandt_breast_v2.R")

```{r fig.height=6, fig.cap="Time depending on simulation setting (second batch). N is the number of patients and NG the number of genes in the pathway."}
plot_time
```

# Discussion

Overall :

- Global Test is fast compared to other approach. It is not completely clear to me why because it also relies on permutation and is similar to Wald test. There is probably a clever implementation ?
- In terms of power Wald test is the best option, closely followed by Global test
- sGBJ is slow when the number of genes rise. This is due to permutation for variance matrix estimation that might be reduced to improve computation time.
- sGBJ performance is similar to most approach except when the number of genes increases.

@@

- estimate one beta by gene using cox model
- sum the wald test statistics

@@

sGBJ is similar to Wald and Global test in the sense that it is based on one cox model by feature. However, it takes into account the covariance of genes in its formula so it should perform better when beta are correlated.
